\section{Introduction}
\label{sec:intro}

Large language models (LLMs) with long context windows face a critical memory bottleneck: the key-value (KV) cache grows linearly with sequence length, consuming $2 \times n \times H \times L \times d \times b$ bits for $n$ tokens, $H$ KV heads, $L$ layers, head dimension $d$, and $b$ bits per element.
For an 8B-parameter model at 128K context in FP16, the KV cache alone requires $\sim$32~GB---often exceeding the model weights.

Two compression paradigms dominate the literature: \emph{quantization} (reducing bits per entry) and \emph{eviction} (reducing the number of entries). Quantization methods such as KIVI~\citep{kivi2024}, KVQuant~\citep{kvquant2024}, and TurboQuant~\citep{turboquant2026} achieve 2--4$\times$ compression, while eviction methods like H\textsubscript{2}O~\citep{h2o2023}, StreamingLLM~\citep{streamingllm2024}, and SnapKV~\citep{snapkv2024} reduce token count by 2--5$\times$. These approaches are typically evaluated using perplexity (PPL) on language modeling benchmarks, with the implicit assumption that low PPL degradation implies preserved model quality.

\paragraph{This paper challenges that assumption.} Through a systematic empirical study across five models (7B--70B parameters, three GQA configurations), we uncover three findings that revise the current understanding of KV cache compression:

\begin{enumerate}
    \item \textbf{Quantization safety is architecture-dependent} (\S\ref{sec:exp-quant}, \S\ref{sec:exp-qwen}). 4-bit KV quantization is NIAH-safe on GQA 4:1 models (Llama, Mistral) but catastrophically fails on GQA 7:1 models (Qwen: NIAH 0\% at every format $\leq$5-bit, with the bit cliff between 5 and 8 bits). Asymmetric analysis reveals \textbf{K-quantization}---not V---as the sole root cause, because K errors enter the softmax nonlinearity where they are amplified across all shared query heads.

    \item \textbf{Eviction has a fundamental causal limitation} (\S\ref{sec:exp-niah}). Four scoring strategies---position-based (StreamingLLM), attention-based (H\textsubscript{2}O), and K-statistics-based (Expected Attention, single- and multi-layer)---all fail at high eviction rates. At 85\% eviction, Expected Attention achieves 0\% NIAH, worse than StreamingLLM's 20\%. This is not an algorithmic failure but a \emph{causal} one: eviction decisions are made at prefill time without access to future queries.

    \item \textbf{Perplexity is a misleading proxy for eviction quality} (\S\ref{sec:exp-metric-gap}). At 50\% eviction, PPL \emph{improves} ($-$0.09\%) while retrieval accuracy drops from 100\% to 60\%---a dangerous false positive. Quantization produces the opposite signature: PPL degrades while retrieval holds. This orthogonality means neither metric alone characterizes compression quality.
\end{enumerate}

\paragraph{Contributions.}
\begin{itemize}
    \item The first systematic \textbf{cross-model, cross-architecture} evaluation of KV cache compression across GQA ratios (4:1, 7:1, 8:1) and model sizes (7B--70B), revealing that GQA ratio is the primary determinant of quantization safety.
    \item A \textbf{GQA-aware asymmetric quantization rule}: K precision must increase with GQA ratio (K$\geq$8-bit for $n_\text{kv}\leq$4), while V is universally safe at 4-bit. Combined with outlier layer detection, this provides an immediately deployable configuration for any model.
    \item Empirical demonstration and theoretical explanation of the \textbf{causal limitation of eviction}, unifying the failure of H\textsubscript{2}O and StreamingLLM as instances of the same information-theoretic constraint.
    \item Proof that K and V quantization errors are \textbf{additive} ($<$0.1\% prediction error), reducing calibration cost from $O(n^2)$ to $O(2n)$ for any asymmetric configuration.
    \item A recommendation for \textbf{dual-metric reporting} (PPL + retrieval task) as the minimum standard for KV cache compression evaluation, supported by our metric gap analysis.
    \item All findings are \textbf{immediately deployable} in llama.cpp---the only mainstream inference framework supporting asymmetric K/V quantization types---without any code modification.
\end{itemize}
