% Paper 2 Conclusion
\section{Conclusion}
\label{sec:conclusion}

We presented PCA-Quant, a principled framework for adaptive KV cache quantization that allocates bits optimally across dimensions via water-filling on the PCA eigenspectrum. Our main findings:

\begin{enumerate}
    \item \textbf{Water-filling provides the theoretical framework; PCA rotation is the practical mechanism.} The classical water-filling result $b_j^* = \bar{b} + \frac{1}{2}\log_2(\lambda_j / \bar{\lambda}_g)$ explains \emph{why} PCA rotation helps: decorrelation equalizes per-block variance, reducing quantization distortion. PCA maximizes coding gain over all orthogonal transforms, including the Hadamard rotation used by TurboQuant. However, adaptive bit allocation across tiers is limited by the coarse granularity of available quantizers (the 4-to-8-bit gap is too large, and sub-4-bit tiers are catastrophic). In practice, PCA rotation with uniform quantization captures the dominant benefit. The overhead is negligible: one-time calibration in $<$1 second, 1 MB storage, $\sim$3\% runtime cost (eliminated via fused query rotation).

    \item \textbf{Coding gain concentrates on outlier layers.} Empirical eigenspectrum analysis reveals that PCA-Quant's benefit is not uniform: outlier layers (2--3 per model) achieve $+2$--$3$ bit effective precision gain, while normal layers gain $+0.4$--$0.5$ bits. This concentration is advantageous---PCA-Quant precisely targets the layers responsible for catastrophic quantization failure in GQA-heavy architectures.

    \item \textbf{Theory matches experiment.} PCA-Quant reduces PPL degradation by 63\% on TinyLlama-1.1B (5 diverse texts, 4/5 improved) and 85\% on Qwen2.5-3B, at the same 4-bit memory budget. The benefit scales with eigenspectral non-uniformity, as predicted by the coding gain theory. The dominant practical mechanism is PCA decorrelation (improving per-block quantization efficiency), rather than adaptive bit allocation---the coarse granularity of available quantization tiers (4-bit to 8-bit) prevents the continuous water-filling optimum from being realized.

    \item \textbf{V cache is near-isotropic.} V cache AM/GM $\approx 1.4$ across all models and layers, confirming that adaptive allocation is needed only for K. This halves implementation complexity.

    \item \textbf{Compression can improve quality.} Attention merging at $2\times$ compression improves QuALITY accuracy by $+2.2\%$ over baseline---a ``compression as regularization'' effect followed by a sharp cliff between $2\times$ and $5\times$. This non-monotonic pattern challenges the assumption that compression necessarily degrades quality and motivates quality-curve evaluation.
\end{enumerate}

\paragraph{Limitations.}
Block-32 quantization in PCA space introduces a scale mismatch artifact: dimensions with vastly different variances share a block scale factor, causing NIAH to drop from 100\% to 80\% on Qwen2.5-3B (Section~\ref{sec:block-mismatch}). PCA rotation without quantization preserves NIAH perfectly, confirming the artifact is due to the block-quantizer interaction, not the rotation. Adaptive bit allocation via water-filling is theoretically optimal but impractical with current quantizer tiers: sub-4-bit tiers (2-bit) are catastrophic, while the 4-to-8-bit gap is too coarse for meaningful allocation. The prototype uses simulated quantization in Python; a production implementation in llama.cpp is needed to measure actual inference speedup.

\paragraph{Future work.}
Three directions are immediate: (1) per-dimension quantization in PCA space (each component quantized with its own scale along the token axis), which would resolve the block-32 mismatch and restore NIAH to 100\%; (2) production implementation in llama.cpp with the fused query rotation optimization; and (3) compositional compression experiments combining PCA-Quant with attention merging to validate the predicted additivity of orthogonal compression axes.
