\section{Related Work}
\label{sec:related}

\subsection{KV Cache Quantization}

\paragraph{Uniform quantization.}
KIVI~\citep{kivi2024} applies 2-bit quantization to both Keys and Values, achieving $\sim$8$\times$ compression with 1--3\% PPL degradation.
KVQuant~\citep{kvquant2024} uses per-channel quantization with outlier handling.
Both apply uniform bit allocation across K and V; neither examines the interaction between GQA ratio and quantization safety.

\paragraph{Rotation-based quantization.}
TurboQuant~\citep{turboquant2026} (Google) applies Hadamard rotation to make entries approximately i.i.d.\ Gaussian before Lloyd-Max scalar quantization, approaching Shannon rate-distortion limits.
Community reproductions have found that the paper's recommended QJL rotation for Keys actually \emph{hurts} due to softmax-amplified variance, with plain MSE quantization outperforming QJL---consistent with our finding that K-quantization errors are disproportionately amplified.

\paragraph{Asymmetric K/V allocation.}
KVTuner~\citep{kvtuner2025} studies per-layer, per-head sensitivity, finding Keys $\sim$2$\times$ more sensitive than Values.
Our work reveals the \emph{architectural} mechanism: the sensitivity gap is mediated by GQA ratio ($n_q / n_\text{kv}$), because K errors enter the softmax and are amplified across all sharing query heads.
At GQA 7:1 (Qwen), 4-bit K is catastrophic; at GQA 4:1 (Llama, Mistral), 4-bit K is safe.
This architecture-dependent failure mode has not been reported in prior work.

\subsection{KV Cache Eviction}

\paragraph{Attention-based eviction.}
H\textsubscript{2}O~\citep{h2o2023} retains ``heavy hitter'' tokens with highest cumulative attention scores, achieving $\sim$5$\times$ compression with moderate PPL impact.
The original paper evaluates primarily on perplexity.

\paragraph{Position-based eviction.}
StreamingLLM~\citep{streamingllm2024} retains attention sink tokens (first few tokens) plus a sliding window of recent tokens, discarding middle tokens regardless of importance.

\paragraph{Query-aware methods.}
SnapKV~\citep{snapkv2024} uses the instruction/query portion of the prompt as an observation window to identify important tokens before generation.
PyramidKV~\citep{pyramidkv2024} allocates layer-wise budgets based on attention pattern structure.
Both are NIAH-safe \emph{when the query is present at prefill time}, but fail for pre-cached contexts or multi-turn scenarios.

\paragraph{Learned eviction.}
LookaheadKV~\citep{lookaheadkv2026} trains learnable pseudo-query vectors to approximate future attention without seeing the actual query.
ForesightKV~\citep{foresightkv2026} distills oracle future attention into a lightweight scorer via supervised learning and RL.
Both require fine-tuning and are not post-training compatible.

\paragraph{The causal limitation.}
We demonstrate that \emph{no} retrospective scoring method (H\textsubscript{2}O, StreamingLLM, or any variant) can preserve retrieval quality under aggressive eviction, because eviction decisions at prefill time cannot anticipate future queries.
This has been formalized independently: key vectors carry only 0.12 bits of mutual information with future attention patterns~\citep{limits-scoring-2026}, and attention predictability is fundamentally task-dependent~\citep{tappa2025}.

\subsection{Evaluation Methodology}

\paragraph{Perplexity as proxy.}
The vast majority of KV cache compression papers use perplexity as the primary (often sole) evaluation metric.
Notable exceptions include SnapKV (NIAH evaluation), PyramidKV (RULER benchmark), and KVzip (multi-query QA).

\paragraph{The metric gap.}
We show that perplexity and retrieval accuracy respond \emph{oppositely} to eviction: PPL can improve while retrieval collapses.
This mirrors the classical PSNR--VMAF gap in video compression~\citep{vmaf2016}, where pixel-level distortion metrics fail to predict perceptual quality.
The implication is that any paper reporting only PPL for eviction-based compression may be unknowingly presenting a misleading picture of quality.

\subsection{Information-Theoretic Perspectives}

Expected Attention~\citep{expected-attn-2025} models future queries as Gaussian-distributed and computes closed-form expected attention scores, achieving NIAH-safe eviction at up to 125K context without training.
KVzip~\citep{kvzip2025} sidesteps the causal limitation entirely by scoring tokens based on reconstruction difficulty---tokens that are information-theoretically irreplaceable are retained regardless of query.
These training-free approaches suggest that the causal limitation can be partially addressed through statistical modeling, though not to the compression ratios achievable by quantization.
