\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{positioning, calc, arrows.meta}

\graphicspath{{figures/}}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{remark}{Remark}

\title{Beyond Perplexity: Why KV Cache Compression Needs Dual-Metric Evaluation}

\author{
  Anonymous Authors
}

\date{}

\begin{document}
\maketitle

\begin{abstract}
We conduct a systematic empirical study of KV cache compression across five LLMs (Llama-3.1-8B, Llama-3.3-70B, Mistral-7B, Qwen2.5-7B, Ministral-8B), revealing three findings that challenge prevailing assumptions.
First, \textbf{quantization safety is architecture-dependent}: 4-bit KV quantization is lossless on GQA 4:1 models (Llama, Mistral) but catastrophically fails on GQA 7:1 models (Qwen: NIAH 0\% at every format $\leq$5-bit, with the bit cliff between 5 and 8 bits), with K-quantization---not V---as the root cause.
Second, \textbf{eviction has a fundamental causal limitation}: four scoring strategies---position-based (StreamingLLM), attention-based (H\textsubscript{2}O), and K-statistics-based (Expected Attention, single- and multi-layer)---all fail at high eviction rates, with Expected Attention achieving 0\% NIAH at 85\% eviction. This limitation is information-theoretic, not algorithmic: eviction decisions at prefill time cannot anticipate future queries.
Third, \textbf{perplexity is an unreliable proxy}: at 50\% eviction, PPL \emph{improves} ($-$0.09\%) while NIAH drops from 100\% to 60\%, creating a dangerous false positive for researchers using PPL alone.
We show that K and V quantization errors are \textbf{additive} (predicted +3.13\% vs.\ measured +3.14\%), reducing calibration cost from $O(n^2)$ to $O(2n)$ configurations.
We propose GQA-aware asymmetric quantization (K$\geq$8-bit for high GQA ratios, V safe at 4-bit universally) and recommend dual-metric reporting (PPL + retrieval) as the minimum standard for KV cache compression evaluation.
Our findings are immediately deployable in llama.cpp, the only inference framework supporting asymmetric K/V types.
\end{abstract}

\input{sections/introduction}
\input{sections/related}
\input{sections/method}
\input{sections/theory}
\input{sections/metric_gap}
\input{sections/experiments}  % Tables with TBD placeholders
\input{sections/conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
